---
title: "Using the IScalc package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the IScalc package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Below are some practical example showing how one can use the `IScalc` function to 
calculate different insulin sensitivity indexes (ISI).

The functions included in this package expect data to be in vectors, which makes 
them to integrate with a [`tidyverse`](http://tidyverse.org/) workflow. Below you will also find examples
in case your data is not in a observation-per-row format, or what is known as [*tidy*](http://vita.had.co.nz/papers/tidy-data.html).


## Loading the package

`IScalc` will load like any standard `R` package:
```{r message = FALSE, warning = FALSE}
library(gmi)

# dplyr will make it easy to apply the functions to whole datasets.
library(dplyr)
library(tidyr)
```



## Calculating ISI with tidy data

If your data looks like the block below, continue reading this section, otherwise move on
to the next section.

```{r echo = FALSE}
data("OGTT_long")
head(OGTT_long)
```

The data above has one observation per row, which is the ideal format to apply the `IScalc` functions quickly.

As a first example I calculate the revised QUICKI for each subject in the included sample dataset `OGTT_long`. The `ISI_fasting` function will compute an index from fasting values; `glu` (glucose), `ins` (insulin) and `index` must be provided. Additionally, for the revised QUICKI index you will need to provide the `nefa` argument.

```{r}
# OGTT_long %>%
#   filter(timePoint == -60) %>% # keep only fasting values
#   group_by(ID) %>% # calculate index for each subject
#   summarise(InsuSen = ISI_fasting(glu = glucose_mgdl, ins = insulin_microU, 
#                                   nefa = nefa_umolL, index = "Revised QUICKI"))
```

Since our sample dataset also includes values from a 2-hour, 5-sample oral glucose tolerance test (OGTT), we might also go ahead and calculate ISI based on that. I'll demonstrate this using for the NEFA-ISI index:

```{r}
# OGTT_long %>%
#   filter(timePoint %in% c(0, 60, 120)) %>% # select only the relevant timepoints
#   arrange(timePoint) %>% # make sure the timepoints are correctly ordered
#   group_by(ID) %>% # calculate index for each subject
#   summarise(InsuSen = ISI_OGTT(ins = insulin_pmoll, nefa = nefa_umolL, 
#                                bmi = bmi[1], index = "NEFA-ISI")) # bmi[1] ensures only one value 
                                                                  # is passed on to the function, 
                                                                  # instead of a vector of 
                                                                  # repeating values.
```

And so, as I previously demonstrated for fasting values, we obtain a data.frame with a column `InsuSen` (you can use whatever name you wish) with a NEFA-ISI value for each subject.

Below another example involving a little more data wrangling to calculate the Stumvoll-ISI index:

```{r}
# OGTT_long %>%
#   gather(parameter, value, c(glucose_mgdl:insulin_microU)) %>% # gather all the columns into a single one
#   filter((timePoint == 90 & parameter == "glucose_mgdl") | (timePoint == 120 & parameter == "insulin_microU")) %>% # select only the relevant timepoints
#   group_by(ID) %>% # calculate index for each subject
#   summarise(InsuSen = ISI_OGTT(glu = value[parameter == "glucose_mgdl"],
#                                ins = value[parameter == "insulin_microU"], 
#                                bmi = bmi[1], index = "Stumvoll-ISI"))
```



## Dataset in subject-per-row format

In case your data is not tidy with one observation per row, we'll have to use a bit more `dplyr` code to rearrange it. Below is an example dataset:

```{r}
#head(OGTT_untidy)
```

**TODO**
